Trauma treatment delay

### 2023-09-15 manuscriptlongversion.Rmd
Introduction

---
title: "Trauma treatment delay"
author: "Anton Wasielewski"
subtitle: A retrospective cohort study on the factors affecting the delays in trauma
  care
output:
  html_document:
    df_print: paged
  pdf_document: default
  word_document: default
csl: bmcemerg.csl
bibliography: bibliography.bib
---

@@

Trauma systems are infrastructures that exist to provide and optimize care for injured patients starting with injury recognition and triage, transport to appropriate trauma center, inpatient care and outpatient follow-up. Beyond the clinical side, it works with outreach, education and advocacy, data collection through registries, research, funding, and disaster preparedness and response. A comprehensive and functioning trauma system requires strong leadership and engagement at the trauma center, regional and national level [@connolly_woo_lampron_perry_2018]. This system is crucial to provide care for trauma patients, both in reducing morbidity and mortality in this patient group. Earlier studies from Sweden have shown that treating severely injured patients at one of these establishments is associated with a 41% lower adjusted 30-day mortality rate compared to being treated at a non trauma care center due to them being more capable of treating these patients, with patients with higher ISS scores having an even higher protective effect with \>90% for the most critically injured patients [@Candefjord_Asker_Caragounis_2020]. Other studies have shown similar results, showing that treating injured patients at trauma centers is associated with a 15-30% decrease in mortality [@Moran_Lecky_Bouamra_Lawrence_Edwards_Woodford_Willett_Coats_2018][@elso_Tepas_Langland-Orban_Pracht_Papa_Lottenberg_Flint_2006][@MacKenzie_Rivara_Jurkovich_Nathens_Frey_Egleston_Salkever_Scharfstein_2006][@Haas_Stukel_Gomez_Zagorski_De_Mestral_Sharma_Rubenfeld_Nathens_2012]. During the last three decades, the introduction of trauma systems has lowered the incidence of preventable death from approximately 25% to less than 5%. This is attributed to the improvements in care for acute brain injuries and bleeding control. The incidence of late death because of sepsis and multiple organ failure, possibly a result of better and earlier resuscitation [@Asensio_Trunkey_2008].

Trauma teams are multidisciplinary and operate in these centers. In Sweden, they consist of individuals from the specialities of surgery, intensive care, orthopedics, nursing and support staff and are lead by a team leader. Their aim is to rapidly assess and stabilise the patient, prioritise their injuries and arrange for site of definitive care. Emergency nurses are often the personnel activating the trauma code using anatomical and physiological criteria, mechanism of injury and information obtained from emergency service personnel. Once in the trauma room, the team works through ABCDE to treat and stabilise the patient following the protocol present at that specific hospital [@Hedberg_Falkén_Wihlke_Jansson_Hultgren_Eriksson_Beckman_Haupt_Kjörholt_Sandersjöö_et_al._2020].

Trauma patients brought to trauma centers may receive immediate or delayed treatment depending on what protocols are present at the facility and how they are being followed [@ryb2012delayed]. Among the preventable or possibly preventable deaths in trauma patients, a group determined by the Combined Trauma Death Review Committee, delay in treatment has been identified as a major error contributing to death, found in up to 52.9% of patients in said group [@Teixeira_Inaba_Hadjizacharia_Brown_Salim_Rhee_Browder_Noguchi_Demetriades_2007]. In the case of a DTTA it can often be attributed to insensitivity in triage protocol and failure to recognize injury patterns [@Hoyt_Hollingsworth-Fridlund_Winchell_Simons_Holbrook_Fortiage_1994]. Studies have also shown that different factors such as age over 55, non-white ethnicity, blunt assault among others was linked to a higher risk of DTTA [@ryb2012delayed].

>>>>

---
title: "Trauma treatment delay"
author: "Anton Wasielewski"
subtitle: A retrospective cohort study on the factors affecting the delays in trauma
  care
output:
  word_document: default
  html_document:
    df_print: paged
  pdf_document: default
bibliography: bibliography.bib
---

@@

Trauma systems are infrastructures that exist to provide and optimize care for injured patients starting with injury recognition and triage, transport to appropriate trauma center, inpatient care and outpatient follow-up. Beyond the clinical side, it works with outreach, education and advocacy, data collection through registries, research, funding, and disaster preparedness and response. A comprehensive and functioning trauma system requires strong leadership and engagement at the trauma center, regional and national level [@connolly_woo_lampron_perry_2018]. This system is crucial to provide care for trauma patients, both in reducing morbidity and mortality in this patient group. Earlier studies from Sweden have shown that treating severely injured patients at one of these establishments is associated with a 41% lower adjusted 30-day mortality rate compared to being treated at a non trauma care center due to them being more capable of treating these patients, with patients with higher ISS scores having an even higher protective effect with \>90% for the most critically injured patients [@Candefjord_Asker_Caragounis_2020]. Other studies have shown similar results, showing that treating injured patients at trauma centers is associated with a 15-30% decrease in mortality [@Moran_Lecky_Bouamra_Lawrence_Edwards_Woodford_Willett_Coats_2018][@Celso_Tepas_Langland-Orban_Pracht_Papa_Lottenberg_Flint_2006][@MacKenzie_Rivara_Jurkovich_Nathens_Frey_Egleston_Salkever_Scharfstein_2006][@Haas_Stukel_Gomez_Zagorski_De_Mestral_Sharma_Rubenfeld_Nathens_2012]. During the last three decades, the introduction of trauma systems has lowered the incidence of preventable death from approximately 25% to less than 5%. This is attributed to the improvements in care for acute brain injuries and bleeding control. The incidence of late death because of sepsis and multiple organ failure, possibly a result of better and earlier resuscitation [@Asensio_Trunkey_2008].

Trauma teams are multidisciplinary and operate in these centers. They play a pivotal role in the initial care of the trauma patient, as they are primary care givers in the critical stage of trauma. In Sweden, they consist of individuals from the specialities of surgery, intensive care, orthopedics, nursing and support staff and are lead by a team leader. Their aim is to rapidly assess and stabilise the patient, prioritise their injuries and arrange for site of definitive care. Emergency nurses are often the personnel activating the trauma code using anatomical and physiological criteria, mechanism of injury and information obtained from emergency service personnel. Once in the trauma room, the team works through ABCDE, a systematic and structured method to assess and manage critically ill patients. It is designed to handle the most urgent problems first, such as airways and breathing, and then go through the rest of the body in a structured way so to not miss anything. [@Hedberg_Falkén_Wihlke_Jansson_Hultgren_Eriksson_Beckman_Haupt_Kjörholt_Sandersjöö_et_al._2020].

**Delay in treatment**

Trauma patients brought to trauma centers may receive immediate or delayed treatment depending on what protocols are present at the facility and how they are being followed [@ryb2012delayed]. A cornerstone of trauma quality improvement programs is multidisciplinary mortality and morbidity review to identify opportunities for improvement and implement corrective actions. This would be a meeting where doctors from different specialities sit down and go through cases where possible OFIs were identified. Examples of opportunities for improvement may include lack of resources and management errors. Among the preventable or possibly preventable deaths in trauma patients, a group determined by the Combined Trauma Death Review Committee, delay in treatment has been identified as a major error contributing to death, found in up to 52.9% of patients in said group [@Teixeira_Inaba_Hadjizacharia_Brown_Salim_Rhee_Browder_Noguchi_Demetriades_2007].
